Towards a miniKanren with fair search strategies

### 2019-05-22 Paper/paper.tex
dan

\newcommand{\BFSser}[0]{BFS$_\textrm{ser}$}

\newcommand{\veryrecursiveo}[0]{\texttt{very-recursive$^o$}}
\newcommand{\appendo}[0]{\texttt{append$^o$}}
\newcommand{\reverso}[0]{\texttt{revers$^o$}}

\newtheorem{defn}{Definition}[section]

@@

To summarize our contribution, we
\begin{itemize}
	\item propose and implement balanced interleaving depth-first search 
	(\DFSbi{}), a new search strategy with almost-fair \disj{}.
	\item propose and implement fair depth-first search (\DFSf{}), a 
	new search strategy with fair \disj{}.
	\item implement in a new way breath-first search (\BFSopt), 
        a search strategy 
	with fair \disj{} and fair \conj{}. And we prove formally that our 
\BFSopt{} implementation is semantically equivalent to the one 
by \citet{seres1999algebra}, however, Our code 
runs faster in all benchmarks and we believe it is simpler.
\end{itemize}

@@

In this section, we define fair \disj{}, almost-fair \disj{} and fair \conj{}. 
Before going further into fairness, we give a short review of the terms:
\emph{state}, search \emph{space}, \emph{goal}.
A \emph{state} is a collection of constraints. (Here, we restrict 
constraints to unification constraints.) Every answer corresponds to a 
state. A \emph{space} is a collection of states. And a \emph{goal} is a 
function from a state to a space.

@@

Given the following program, it is natural to expect lists of each letter to
constitute $1/4$ in the answer. Search strategies with unfair \conj{} (e.g. 
\DFSi, \DFSbi, \DFSf), however, results in many more lists of \texttt{a}s than 
lists of other letters. And some letters are rarely seen. The situation would 
be exacerbated if \conde{} were to contain more clauses. Although 
DFS$_\textrm{i}$'s \disj{} is unfair in general, it is fair when there is no 
call to a relational definition in \conde{} clauses (including this case).

@@

A more interesting situation is when the first conjunct produces an unbounded
number of states. Consider the following example, a naive specification of fair \conj{} 
might require search strategies to produce all sorts of singleton lists, but no 
longer ones, which makes the strategies incomplete. A search strategy is 
\emph{complete} if and only if it can determine all the answers within finite 
time, otherwise, it is \emph{incomplete}.

@@

\BFSser{} puts states of the same cost in the same bag, and allocates
resources carefully among search spaces related to different bags such
that answers are produced in increasing order of cost. The \emph{cost}
of an answer is its depth in the search tree (i.e., the number of
calls to relations required to find them) \citep{seres1999algebra}. In
the following example, every answer is a list of list of symbols,
where inner lists are the same. Here the cost of each answer is equal
to the length of its inner lists plus the length of its outer list.

@@

In this section, we review the implementation of interleaving depth-first 
search (\DFSi). We focus on parts that are relevant to this paper. TRS2 
\citep{friedman_reasoned_2018} provides a comprehensive description of the 
whole miniKanren implementation.

@@

\NOTE[DPF]{Why does the pair? case precede the null? case? Is it simply more efficient? It shouldn't be?}

@@

The new type is a pair whose \texttt{car} is a list of answers (the bag), and 
whose \texttt{cdr} is either a \texttt{\#{}f} or a thunk returning a search 
space. A falsy \texttt{cdr} means the search space is obviously finite.

@@

Functions implementing impure features are in Fig.~\ref{BFS-opt-cont}. The 
first function, \texttt{elim}, takes a space \texttt{s$^\infty$} and two 
continuations \texttt{ks} and \texttt{kf}. When \texttt{s$^\infty$} contains 
some 
answers, \texttt{ks} is called with the first answer and the rest space. 
Otherwise, \texttt{kf} is called with no argument. Here `s' and `f' means 
`succeed' and `fail' respectively. This function is like an eliminator of 
search space, hence the name. The remaining functions do the same thing as 
before.

@@

In this section, we compare the pure subset of \BFSopt{} with the BFS 
found in \citet{seres1999algebra}. We focus on the pure subset because 
their system is pure. Their system represents search spaces with streams of 
lists of answers, where each list is a bag.

To compare efficiency, we translate their Haskell code into Racket (See 
supplements for the translated code). The translation is direct 
due to the similarity in both logic programming systems and search space 
representations. The translated code is longer and slower than BFSopt
implementation. Details about 
difference in efficiency are in section 7.

We prove in Coq that \BFSser{} and \BFSopt{} are semantically equivalent, i.e. \texttt{(run n ? g)} 
produces the same result (See supplements for the formal proof).

@@

\begin{table}
	\begin{tabular}{|c|c|c|c|c|c|c|}
		\hline 
		benchmark & size & DFS$_\textrm{i}$ & DFS$_\textrm{bi}$ & DFS$_\textrm{f}$ & BFS$_\textrm{opt}$ & BFS$_\textrm{ser}$  
		\\
		\hline
		\veryrecursiveo{} & 100000 &  579 &  793 & 2131 & 1438 & 3617 \\
		& 200000 & 1283 & 1610 & 3602 & 2803 & 4212 \\
		& 300000 & 2160 & 2836 &    - & 6137 &    - \\
		\hline 
		\appendo{}  & 100 &  31 &  41 &  42 &  31 &  68 \\ 
		& 200 & 224 & 222 & 221 & 226 & 218 \\ 
		& 300 & 617 & 634 & 593 & 631 & 622 \\ 
		\hline 
		\reverso{} & 10 &   5 &   3 &   3 &     38 &     85 \\ 
		& 20 & 107 &  98 &  51 &   4862 &   5844 \\
		& 30 & 446 & 442 & 485 & 123288 & 132159 \\ 
		\hline
		quine-1 & 1 &  71 &  44 & 69 & - & - \\ 
		& 2 & 127 & 142 & 95 & - & - \\ 
		& 3 & 114 & 114 & 93 & - & - \\ 
		\hline
		quine-2 & 1 & 147 & 112 &  56 & - & - \\ 
		& 2 & 161 & 123 & 101 & - & - \\ 
		& 3 & 289 & 189 & 104 & - & - \\ 
		\hline 
		'(I love you)-1 &  99 & 56 & 15 & 22 &  74 & 165 \\ 
		& 198 & 53 & 72 & 55 &  47 &  74 \\
		& 297 & 72 & 90 & 44 & 181 & 365 \\ 
		\hline
		'(I love you)-2 &  99 & 242 &  61 & 16 &  66 &  99 \\ 
		& 198 & 445 & 110 & 60 &  42 &  64 \\
		& 297 & 476 & 146 & 49 & 186 & 322 \\ 
		\hline 
	\end{tabular}
	\caption{The results of a quantitative evaluation: running times of 
	benchmarks 
		in milliseconds}
	\label{compare-efficiency}
\end{table}

Here we compare the efficiency of search strategies. A concise 
description is in Table~\ref{compare-efficiency}. A hyphen means ``running out of 
memory.'' The first two benchmarks are taken from 
TRS2. \reverso{} is from 
\citet{rozplokhas2018improving}. The next two benchmarks 
about quine are modified from a similar test case in \citet{byrd2017unified}. 
The modifications are made 
to circumvent the need for symbolic constraints (e.g. $\neq$, 
\texttt{absent$^o$}). Our version generates de 
Bruijnized expressions and prevent closures getting into list. The two 
benchmarks differ in the \conde{} clause order of their relational 
interpreters. 
The last two 
benchmarks are about synthesizing expressions that evaluate to \texttt{'(I love 
you)}. This benchmark is from \citet{byrd2017unified}. Again, the 
sibling benchmarks differ in the \conde{} clause order of their relational 
interpreters. The first one 
has elimination rules (i.e. application, \texttt{car}, and \texttt{cdr}) at the 
end, while the other has them at the beginning. We conjecture that DFS$_\textrm{i}$ would 
perform badly in the second case because elimination rules complicate the 
problem when synthesizing (i.e., our evaluation supports our conjecture.)

In general, only DFS$_\textrm{i}$ and DFS$_\textrm{bi}$ constantly perform well. DFS$_\textrm{f}$ is just as 
efficient in all benchmarks but \veryrecursiveo{}. Both BFS have obvious 
overhead in many cases. Among the three variants of DFS (they all have unfair 
\conj{}), DFS$_\textrm{f}$ is most resistant to clause permutation, followed by DFS$_\textrm{bi}$ then 
DFS$_\textrm{i}$. Among the two implementation of BFS, ours constantly performs as well or 
better. Interestingly, every strategy with fair \disj{} suffers in 
\veryrecursiveo{} and DFS$_\textrm{f}$ performs well elsewhere. 
Therefore, this 
benchmark might be a special case. Fair \conj{} imposes considerable overhead 
constantly except in \appendo{}. The reason might be that strategies with 
fair \conj{} tend to keep more intermediate answers in the memory.

@@

Yang \citet{yang2010adventures} points out a disjunct complex would be `fair' 
if it were a full and balanced tree .

Seres et al \citet{seres1999algebra} also describe a breadth-first search 
strategy. We prove \BFSser{} is semantically equivalent to \BFSopt. But our codeis shorter and performs better in comparison with a straightforward 
translation of their Haskell code.

@@

Our quantitative evaluation shows that DFS$_\textrm{bi}$ and DFS$_\textrm{f}$ are competitive 
alternatives to DFS$_\textrm{i}$, the current search strategy, and that 
\BFSopt is less practical than other search strategies.
\NOTE[DPF]{what}
\NOTE[LKC]{fixed}

We prove that \BFSopt{} is equivalent to \BFSser in \citet{seres1999algebra}. 
Our optimized code is shorter and runs faster than a direct translation 
of their Haskell code.

>>>>

\newcommand{\BFSser}[0]{BFS$_\textrm{ser}$}
\newtheorem{defn}{Definition}[section]

@@

To summarize our contribution, we
\begin{itemize}
	\item propose and implement balanced interleaving depth-first search 
	(\DFSbi{}), a new search strategy with almost-fair \disj{}.
	\item propose and implement fair depth-first search (\DFSf{}), a 
	new search strategy with fair \disj{}.
	\item implement in a new way breath-first search (\BFSopt), 
        a search strategy 
	with fair \disj{} and fair \conj{}. And we prove formally that our 
\BFSopt{} implementation is semantically equivalent to \BFSser
however, Our code runs faster in all benchmarks and we believe it is simpler.
\end{itemize}

@@

In this section, we define fair \disj{}, almost-fair \disj{} and fair \conj{}. 
Before going further into fairness, we give a short review of the terms:
\emph{state}, search \emph{space}, \emph{goal}.
A \emph{state} is a collection of constraints. (Here, we restrict 
constraints to unification constraints. Every answer corresponds to a 
state. A space is a collection of states. And a \emph{goal} is a 
function from a state to a space.

% Goals check their input states and outputs possibly extended states. A goal 
% might fail for some inputs, in which case the output search space would be 
% empty.

@@

Given the following program, it is natural to expect lists of each letter to
constitute $1/4$ in the answer. Search strategies with unfair \conj{} (e.g. 
\DFSi, \DFSbi, \DFSf), however, results in many more lists of \texttt{a}s than 
lists of other letters. And some letters are rarely seen. The situation would 
be exacerbated if \conde{} were to contain more clauses.

\NOTE[LKC]{Should I add a footnote?}

``
Although 
DFS$_\textrm{i}$'s \disj{} is unfair in general, it is fair when there is no 
call to a relational definition 
in \conde{} clauses (including this case).

@@

A more interesting situation is when the first conjunct produces an unbounded
number of states. Consider the following example, a naive specification of fair \conj{} 
might require search strategies to produce all sorts of singleton lists, but there
would not be any lists of length two or longer, which makes the strategies incomplete. 
A search strategy is \emph{complete} if and only if it can determine all the answers within finite 
time, otherwise, it is \emph{incomplete}.

\NOTE[DPF]{Does this assume that we never use run*, since in fact, the paper does not
include and uses of run*.  Something should be mentioned around here.}

@@

\BFSser{} puts states of the same cost in the same bag, and allocates
resources carefully among search spaces related to different bags such
that answers are produced in increasing order of cost. The \emph{cost}
of an answer is its depth in the search tree (i.e., the number of
calls to relations required to find them) \citep{seres1999algebra}. In
the following example, every answer is a list of a list of symbols,
where inner lists are the same. Here the cost of each answer is equal
to the length of its inner list plus the length of its outer list.

@@

In this section, we review the implementation of interleaving depth-first 
search (\DFSi). We focus on parts that are relevant to this paper. TRS2,
chapter 10 and the appendix, ``Connecting the wires,'' 
provides a comprehensive description of the 
miniKanren 
implementation but limited to unification constraints ($\equiv$).

@@

\NOTE[DPF]{Why does the pair? case precede the null? case? Is it simply more efficient? It shouldn't be}

@@

The new type is a pair whose \texttt{car} is a list of answers (the bag), and 
whose \texttt{cdr} is either \texttt{\#{}f} or a thunk returning a search 
space. A falsy \texttt{cdr} means the search space is obviously finite.

@@

Functions implementing impure features are in Fig.~\ref{BFS-opt-cont}. The 
first function, \texttt{elim}, takes a space \texttt{s$^\infty$} and two 
continuations \texttt{ks} and \texttt{kf}. When \texttt{s$^\infty$} contains 
some 
answers, \texttt{ks} is called with the first answer and the rest of the space. 
Otherwise, \texttt{kf} is called with no argument. Here `s' and `f' means 
`succeed' and `fail' respectively. This function is like an eliminator of 
search spaces, hence the name. The remaining functions do the same thing as 
before.

@@

In this section, we compare the pure subset of \BFSopt{} with the 
\BFSser{}. We focus on the pure subset because \BFSser{} is pure.
and represents search spaces with streams of 
lists of answers, where each list is a bag.

To compare efficiency, we translate \BFSser{}'s Haskell code into Racket (See supplements for the translated code). The translation is direct 
due to the similarity in both logic programming systems and search space 
representations. The translated code is longer and slower than \BFSopt{}'s implementation. Details about 
differences in efficiency are in the table below.

We use Coq to show that \BFSser{} and \BFSopt{} are 
semantically equivalent, (i.e., \texttt{(run n ? g)} 
produces the same result (See supplements for the formal proof).

@@

\begin{table}
	\begin{tabular}{|c|c|c|c|c|c|c|}
		\hline 
		benchmark & size & DFS$_\textrm{i}$ & DFS$_\textrm{bi}$ & DFS$_\textrm{f}$ & BFS$_\textrm{opt}$ & BFS$_\textrm{ser}$  
		\\
		\hline
		very-recursiveo & 100000 &  579 &  793 & 2131 & 1438 & 3617 \\
		& 200000 & 1283 & 1610 & 3602 & 2803 & 4212 \\
		& 300000 & 2160 & 2836 &    - & 6137 &    - \\
		\hline 
		appendo  & 100 &  31 &  41 &  42 &  31 &  68 \\ 
		& 200 & 224 & 222 & 221 & 226 & 218 \\ 
		& 300 & 617 & 634 & 593 & 631 & 622 \\ 
		\hline 
		reverso & 10 &   5 &   3 &   3 &     38 &     85 \\ 
		& 20 & 107 &  98 &  51 &   4862 &   5844 \\
		& 30 & 446 & 442 & 485 & 123288 & 132159 \\ 
		\hline
		quine-1 & 1 &  71 &  44 & 69 & - & - \\ 
		& 2 & 127 & 142 & 95 & - & - \\ 
		& 3 & 114 & 114 & 93 & - & - \\ 
		\hline
		quine-2 & 1 & 147 & 112 &  56 & - & - \\ 
		& 2 & 161 & 123 & 101 & - & - \\ 
		& 3 & 289 & 189 & 104 & - & - \\ 
		\hline 
		'(I love you)-1 &  99 & 56 & 15 & 22 &  74 & 165 \\ 
		& 198 & 53 & 72 & 55 &  47 &  74 \\
		& 297 & 72 & 90 & 44 & 181 & 365 \\ 
		\hline
		'(I love you)-2 &  99 & 242 &  61 & 16 &  66 &  99 \\ 
		& 198 & 445 & 110 & 60 &  42 &  64 \\
		& 297 & 476 & 146 & 49 & 186 & 322 \\ 
		\hline 
	\end{tabular}
	\caption{The results of a quantitative evaluation: running times of 
	benchmarks 
		in milliseconds}
	\label{compare-efficiency}
\end{table}

Here we compare the efficiency of search strategies. A concise 
description is in Table~\ref{compare-efficiency}. A hyphen means ``running out of 
memory.'' The first two benchmarks are taken from 
TRS2. \texttt{reverso} is from 
\citet{rozplokhas2018improving}. The next two benchmarks 
about quine are modified from a similar test case in \citet{byrd2017unified}. 
The modifications are made 
to circumvent the need for symbolic constraints (e.g. $\neq$, 
\texttt{absent$^o$}). Our version generates de 
Bruijnized expressions and prevents closures from being inside a list. The two 
benchmarks differ in the \conde{} clause order of their relational 
interpreters. 
The last two 
benchmarks are about synthesizing expressions that evaluate to \texttt{'(I love 
you)}. This benchmark is from \citet{byrd2017unified}. Again, the 
sibling benchmarks differ in the \conde{} clause order of their relational 
interpreters. The first one 
has elimination rules (i.e. application, \texttt{car}, and \texttt{cdr}) at the 
end, while the other has them at the beginning. We conjecture that DFS$_\textrm{i}$ would 
perform badly in the second case because elimination rules complicate the 
problem when synthesizing (i.e., our evaluation supports our conjecture.)

\NOTE[DPF]{Below, veryrecursiveo (if possible, should have its o raised.)
One way to do this is to have a different macro.  In the table there are
other occurrences that need their names with a different macro.}

In general, only DFS$_\textrm{i}$ and DFS$_\textrm{bi}$ constantly perform well. DFS$_\textrm{f}$ is just as 
efficient in all benchmarks but \texttt{very-recursiveo}. Both BFS have obvious 
overhead in many cases. Among the three variants of DFS (they all have unfair 
\conj{}), DFS$_\textrm{f}$ is most resistant to clause permutation, followed by DFS$_\textrm{bi}$ then 
DFS$_\textrm{i}$. Among the two implementation of BFS, ours constantly performs as well or 
better. Interestingly, every strategy with fair \disj{} suffers in 
\texttt{very-recursiveo} and DFS$_\textrm{f}$ performs well elsewhere. 
Therefore, this 
benchmark might be a special case. Fair \conj{} imposes considerable overhead 
constantly except in \texttt{appendo}. The reason might be that strategies with 
fair \conj{} tend to keep more intermediate answers in the memory.

@@

\citet{yang2010adventures} points out a disjunct complex would be `fair' 
if it were a full and balanced tree.

\citet{seres1999algebra} also describe a breadth-first search 
strategy. We prove \BFSser{} is semantically equivalent to \BFSopt. But,
The implementation of \BFSopt is shorter and performs better in comparison with a straightforward translation of their Haskell code.

@@

Our quantitative evaluation shows that DFS$_\textrm{bi}$ and DFS$_\textrm{f}$ are competitive 
alternatives to DFS$_\textrm{i}$, the current search strategy, and that 
\BFSopt is less practical than 
\NOTE[DPF]{what?}

We prove that \BFSopt{} is equivalent to \BFSser. 
Our optimized code is shorter and runs faster than a direct translation 
of their Haskell code.

% We acknowledge that we are making large claims about efficiency given
% that we have just a few benchmarks, but our intuition is based on
% sound principles.

\NOTE[DPF]{Finish this paragraph with a clarification and you may
place this anywhere in the conclusion. This may require some consultation with Weixi.}
